---
title: "Global Declining Tree Cover"
author: 'Jillian Hall'
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    theme: yeti
---

```{r setup, include=FALSE}
#loading in all the packages I may need 
library(flexdashboard)
library(readr)
library(dplyr)
library(knitr)
library(tidyr)
library(ggplot2)
library(plotly)
library(leaflet)
library(rnaturalearth)
library(sf)
library(tinytable)

#importing data via URL using readr
#renamed long column names 
URL<-"https://ourworldindata.org/grapher/proportion-of-forest-area-with-long-term-management-plan.csv?v=1&csvType=full&useColumnShortNames=true"
forest_manag<-read_csv(URL)
forest_manag<-forest_manag%>%rename(country=entity,forestmgt='_15_2_1__ag_lnd_frstmgt')

URL2<-"https://ourworldindata.org/grapher/tree-cover-loss.csv?v=1&csvType=full&useColumnShortNames=true"
tree_loss<-read_csv(URL2)
tree_loss<-tree_loss%>%rename(tree_cover_loss=tree_cover_loss_ha__category_total,country=entity)

URL3<-"https://ourworldindata.org/grapher/tree-cover-loss-by-dominant-driver.csv?v=1&csvType=full&useColumnShortNames=true"
loss_driver<-read_csv(URL3)
loss_driver<-loss_driver%>%rename(settlement_infrastructure=tree_cover_loss_ha__category_settlements__and__infrastructure,hard_commodities=tree_cover_loss_ha__category_hard_commodities,natural_disturbances=tree_cover_loss_ha__category_other_natural_disturbances,shifting_cultivation=tree_cover_loss_ha__category_shifting_cultivation,logging=tree_cover_loss_ha__category_logging,permanent_agriculture=tree_cover_loss_ha__category_permanent_agriculture,wildfire=tree_cover_loss_ha__category_wildfire,country=entity)

```

# The Challenge: Global Forest Loss

## Left Column

### Declining Global Tree Cover

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
#filtering global tree cover data for the row "world"
worldtreeloss<-tree_loss%>%filter(country=="World")

#making tree cover loss represent Mha for better visualization
worldtreecover<-worldtreeloss%>%mutate(Ha_millions=tree_cover_loss/1000000)

#finding the total tree cover loss for 2001-2024
worldtreecover<-worldtreecover%>%group_by(country)%>%summarise(total_loss=sum(Ha_millions))
```

Global forests are essential for climate regulation, biodiversity, and human livelihoods, yet they continue to decline worldwide. Although roughly **4.06 billion hectares** of forest remain, nearly one-third of global forest cover has already been lost.
Satellite data show that between 2001 and 2024, the world lost approximately **515 million hectares** of tree cover, an area about half the size of the United States. Understanding where and why this loss occurs is central to achieving **Sustainable Development Goal 15**, which aims to protect and restore terrestrial ecosystems. 

### Figure {.no-padding}

```{r fig.height=13, fig.width=19, message=FALSE, warning=FALSE, paged.print=FALSE}

#adding column for Mha to original data set 
tree_loss<-tree_loss%>%mutate(Ha_millions=tree_cover_loss/1000000)

#filtering data set to remove codes that dont coordinate to a country
#also making total loss of Mha column 
tree_loss1<-tree_loss%>%filter(nchar(code)==3)%>% group_by(country,code)%>%summarise(total_loss=sum(Ha_millions,na.rm=TRUE))%>%ungroup()

#making map that will be the base of the figure 
world <- ne_countries(scale = "medium", returnclass = "sf")

#joining the map with the tree loss data by country code 
map_data<- world %>%left_join(tree_loss1, by = c("iso_a3" = "code"))

#making plot that shows total tree loss from 2001-2024 using geom_sf and saving it as an object 
treeworld<-ggplot(map_data)+geom_sf(aes(fill=total_loss,label=name), color = "grey29") +
  scale_fill_gradient(
    low="darkgreen",
    high= "honeydew",
    name = "Mha",
    na.value = "grey90")+
  labs(
    title = "Tree cover loss is concentrated in a small number of countries, highlighting where \n SDG 15 efforts may have the greatest global impact."
  )+theme_void(base_size =12)
#making an interactive plot 
ggplotly(treeworld)

```

## right column

### Importance of Forests

Forests provide critical ecosystem services, including carbon storage, climate regulation, and resources that support human well-being worldwide. The global value of these services has been estimated at **$16.2 trillion annually**, highlighting the environmental and economic consequences of continued forest loss. (Costanza et al., 2014).

### Lead Contributors {.no-padding}

```{r, message=FALSE, warning=FALSE, paged.print=FALSE}
#selecting country and total loss column from data set 
tree_loss2<-tree_loss1%>%select(country,total_loss)

#renaming columns to make table look more professional
tree_loss2<-tree_loss2%>%rename(Country=country,'Total Loss'=total_loss)

#arranging the total loss in descending order while only taking the top five rows 
tree_loss2 %>% arrange(desc(`Total Loss`))%>%slice(1:5)%>%tt(caption="Countries Driving the Largest Share of Global Forest Loss (2001-2024)")

```

# Why Forests Are Disappearing

## Left column

### Drivers of Tree Cover Loss Through the Years {.no-padding}

```{r fig.height=6, fig.width=5}
#filtering the drivers of tree loss data set for "world" rows
worlddrivers<-loss_driver%>%filter(country=='World')

#making the data set longer putting individual driver columns under one overall driver column 
worlddriverlong<-worlddrivers%>%
  pivot_longer(cols=settlement_infrastructure:wildfire,names_to = "driver",values_to ="ha")

# adding a column that mutates ha to be in the millions
worlddriverlong<-worlddriverlong%>%mutate(ha_mil=ha/1000000)

#reordering drivers to be in acceding order
worlddriverlong<-worlddriverlong%>%mutate(driver=factor(driver,levels = c("permanent_agriculture",
    "wildfire",
    "logging",
    "shifting_cultivation",
    "hard_commodities",
    "settlement_infrastructure",
    "natural_disturbances")))%>%arrange(driver)

# making this into a graph for driver of tree loss per year and loss of Mha and saving it as an object
driverplot<-ggplot(worlddriverlong,aes(x=year,y=ha_mil,fill=driver))+
  geom_area(position='stack',alpha=0.8)+
  labs(y='Mha',
       title='Permanent Agricultural expansion consistently drives \n the majority of tree cover loss.')+
  theme_classic(base_size = 10)
#making the plot interactive through plotly
ggplotly(driverplot)
```

### Temporary vs Permanent {.no-padding}

```{r echo=FALSE, fig.height=5, fig.width=4, message=FALSE, warning=FALSE, paged.print=FALSE}

#using the long data and grouping it by country and driver to find the total Mha 
worlddriverlong<-worlddriverlong%>%group_by(country,driver)%>%summarise(Total=sum(ha_mil))

#ungrouping the long data set and selecting driver and total columns and reordering how the driver column will be presented
worlddriverlong<-worlddriverlong%>%ungroup()%>%select(driver, Total)%>%mutate(driver=factor(driver,levels=c('permanent_agriculture','hard_commodities','settlement_infrastructure','logging','wildfire','shifting_cultivation','natural_disturbances')))%>%arrange(driver)

#making a tiny table of loss per driver but grouping them by temporary or permenant effects and then bolding the group titles 
worlddriverlong %>%
  tt(caption = "Most Tree Cover Loss Results from Permanent Land-Use Change") %>%
  group_tt(i = list(
    "Permanent" = 1,
    "Temporary" = 4
  ))%>%
  style_tt(
    i = c(1, 5),
    bold = TRUE
  )
  
```

## Right column

### Figure {.no-padding}

```{r fig.height=11, fig.width=8}
#making a world drivers total data set grouping by country and finding the total ha per each driver with summaries 
worlddrivertotal<-worlddrivers%>%
  group_by(country)%>%
  summarise(total_cultivation=sum(shifting_cultivation),
            total_logging=sum(logging),
            total_infrastructure=sum(settlement_infrastructure),
            total_wildfire=sum(wildfire),
            total_commodities=sum(hard_commodities),
            total_natural=sum(natural_disturbances),
            total_agri=sum(permanent_agriculture))

#mutating the total columns to present Mha for better visualizations 
worlddrivertotal<-worlddrivertotal%>%mutate(total_logging=total_logging/1000000, total_infrastructure=total_infrastructure/1000000,total_cultivation=total_cultivation/1000000,total_wildfire=total_wildfire/1000000,total_commodities=total_commodities/1000000,total_natural=total_natural/1000000,total_agri=total_agri/1000000)

#renaming the total ha per driver columns to look better when put into a visualization using rename function. 
worlddrivertotal<-worlddrivertotal%>%rename(Logging=total_logging,'Shifting Cultivation'=total_cultivation,Infrastructure=total_infrastructure,'Natural Disturbance'=total_natural,'Hard Commodities'=total_commodities,Wildfire=total_wildfire,'Permanent Agriculture'=total_agri,Scale=country)

#making a column with the overall total of all tree loss ha from drivers using group by each driver and country and summaries for sum of all drivers
worldtotal<-worlddrivertotal%>%
  group_by(Scale,Logging,Infrastructure,`Natural Disturbance`,Wildfire,`Permanent Agriculture`,`Hard Commodities`,`Shifting Cultivation`)%>%
  summarise(Total=sum(Logging,Infrastructure,`Natural Disturbance`,Wildfire,`Permanent Agriculture`,`Hard Commodities`,`Shifting Cultivation`))

#making a pie chart of the contribution of each driver to the total Mha from 2001-2024 using plotly 
plot_ly(worlddriverlong,labels=~driver, values = ~Total, type = 'pie')%>%layout(title='Agriculture and Fire Account for the Majority of Global Tree Cover Loss')

```

### Lead Drivers

Global tree cover loss is driven primarily by **permanent land-use change**, especially agricultural expansion. While wildfire, logging, and shifting cultivation also contribute, these are often **temporary** and may allow forests to regrow. The dominance of permanent conversion highlights a key challenge for SDG 15: preventing deforestation is often more effective than restoring forests after they are lost. 
(Sims et al., 2025)

# Pathways Toward SDG 15

## top

### Progress {.no-padding}

**Percent of global forests under long term sustainable management plans:**

```{r}
#using gauge to show percent of forest under management and setting color to midnight blue to look more uniform 
gauge(58,min=0,max=100,symbol='%',gaugeSectors(success=c(0,100),colors='midnightblue'))
```

This represents a slight increase over recent decades, reflecting growing global adoption of sustainable forest policies.

### Long-Term Managment {.no-padding}

```{r fig.height=13, fig.width=20}
#using filter function to find the total forest under management today or in 2020 when the data ends 
forest2020<-forest_manag%>%filter(year==2020)

#preparing data to be mapped by filtering for codes that dont coordinate to a country and saving it as an new object
forest2020edited<-forest2020%>%filter(nchar(code) == 3)

#setting up the base from which the data will be mapped on 
world <- ne_countries(scale = "medium", returnclass = "sf")

#joining the world and forest2020edited data sets by country codes accounting for non spatial data
map_data2 <- world %>%
  left_join(forest2020edited, by = c("iso_a3" = "code")) %>%
  filter(!sf::st_is_empty(geometry))

# accounting for differences in geometry 
map_data2 <- sf::st_cast(map_data2, "MULTIPOLYGON")

#making map of forest percentage under management and saving it as an object
manageworld<-ggplot(map_data2) +
  geom_sf(aes(fill = forestmgt,label=name), color = "grey29") +
  scale_fill_gradient(
    low="aliceblue",
    high= "midnightblue",
    name="Percent of Forests",
    na.value ="grey90")+
  labs(
    title = "Sustainable Forest Management Is Expanding but Uneven Across Regions")+theme_void(base_size = 10)
#making the map interactive 
ggplotly(manageworld)
```

## column

### Lead Contributors

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

#using filter function to find the total forest under management today or in 2020 when the data ends 
forest2020<-forest_manag%>%filter(year==2020)
#filtering forest 2020 data set previously created for the top five contributors to tree cover loss 
forests2020<-forest2020%>%filter(country%in%c("Russia","Brazil","Canada","United States","Indonesia"))

#making a tiny table selecting only country and forest management percentage and adding a caption 
forests2020%>%select(country,forestmgt)%>%tt(caption="Major Forest-Loss Countries Show Mixed Progress in Sustainable Management.")

```

### What This Means for SDG 15:
Global forest loss remains high, but the drivers are well understood and proven solutions already exist, making SDG 15 a challenge of implementation rather than discovery.

## 2 row {.tabset .tabset-fade}

### Solutions

Because forest loss has multiple causes, effective solutions must combine policy, sustainable management, and community involvement. Key approaches include:

- Reducing agricultural expansion 
- Promoting sustainable forestry 
- Improving wildfire management 
- Strengthening land and resource rights 
- Supporting restoration and conservation planning 

(Sims et al., 2025) (Reek et al., 2026)

### Successes

**Vietnam** reversed decades of forest loss through large-scale restoration programs, increasing forest cover even as agriculture expanded.

**Tanzania** reduced harvesting below forest growth rates by linking sustainable forest management with poverty-reduction policies.

**Guyana** maintained low deforestation through an international partnership with Norway that financially rewards conservation.

**Madagascar** established a protected forest corridor managed by local communities, reducing emissions while supporting livelihoods.

(Boucher et al., 2014)

